
%% == 1 ==

\begin{dhpVerse}{87-88}
\label{dhp-87}\label{dhp-88}
With an image of liberation as the goal\\
the wise abandon darkness and cherish light,\\
leave petty security behind\\
and seek freedom from attachment.\\
To pursue such release is difficult and rare,\\
yet the wise will seek it,\\
detaching themselves from obstructions,\\
purifying heart and mind.
\end{dhpVerse}

\begin{dhpRefl}

The Buddha offers images that illustrate the goal, uplifting and
supporting us in our effort to let go of that which obstructs and
limits us. If we hold too tightly to the images, we may lose
perspective on the here and now element of the journey; instead of
actually doing the practice, we are imagining it. If we fail to give
right emphasis to the goal, we may become lost in the distraction of
sense objects -- agreeable and disagreeable. The pursuit of true
freedom is difficult but consider how much suffering comes if we
don't practice. With wise reflection we find we can endure the dark
and difficult times. When the light returns, we cherish it and
discover how to love truth more fully.

\end{dhpRefl}

%% == 2 ==

\begin{dhpVerse}{169}
\label{dhp-169}
Live your life well in accord with the Way --\\
avoid a life of distraction.\\
A life well-lived leads to contentment,\\
both now and in the future.
\end{dhpVerse}

\begin{dhpRefl}

With a heart of contentment as our foundation, we can tackle the
tasks that confront us. There are times we need to be brave warriors
battling with the forces of delusion to avoid their taking control of
our hearts and minds. At other times we need to be like a parent,
nurturing and caring for the goodness that is already here. Agility
is a great spiritual virtue. Recognising the beauty inherent in a
contented heart, we will naturally be drawn towards it. We only seek
distraction because we don't know contentment. Right practice
liberates energy previously consumed by compulsiveness. This same
energy can also manifest as vitality and enthusiasm.

\end{dhpRefl}

%% == 3 ==

\begin{dhpVerse}{58-59}
\label{dhp-58}\label{dhp-59}
Just as a sweet scented and beautiful lotus\\
can grow from a pile of discarded waste,\\
the radiance of a true disciple of the Buddha\\
outshines dark shadows cast by ignorance.
\end{dhpVerse}

\begin{dhpRefl}

We tend to see the things we don't like about ourselves and others as
obstructions to our happiness. `How radiant, how fulfilled I would
feel, how acceptable I would become if only I didn't have all these
limitations!' But everything we experience can be turned around and
can nourish growth towards that which is inherently, timelessly
beautiful. All the unfair, unwanted, unreasonable stuff of our lives,
all the bits that we resist and reject, make up the swampland of
discarded waste. The beautiful lotuses take root and grow in the
unbeautiful.

\end{dhpRefl}

%% == 4 ==

\begin{dhpVerse}{290}
\label{dhp-290}
It is wisdom\\
that enables letting go\\
of a lesser happiness\\
in pursuit of a happiness\\
which is greater.
\end{dhpVerse}

\begin{dhpRefl}

The filters of our preferences tragically limit our seeing. We want
to let go of that which binds us but often our will fails us. Wise
reflection can support will -- it is will's best friend. Will is not
meant to do it alone. The verse encourages us to reflect on how
letting go of our attachment to a lesser happiness can lead to
gaining a greater happiness. Lost in our attachments, we see only
that which we stand to lose by letting go. Wise thinking means we see
what we stand to lose as well as what we stand to gain. Wise
reflection opens and broadens our vision and enables pursuit of the
goal.

\end{dhpRefl}

%% == 5 ==

\begin{dhpVerse}{227-228}
\label{dhp-227}\label{dhp-228}
Since ancient times it has been the case\\
that those who speak too much are criticised,\\
as are those who speak too little\\
and those who don't speak at all.\\
Everyone in this world is criticised.\\
There never was, nor will there be,\\
nor is there now,\\
anybody who is only blamed or wholly praised.
\end{dhpVerse}

\begin{dhpRefl}

Whatever we do in this life, whatever we say (or don't say), we
cannot avoid being criticised. The Buddha was blamed and criticised
just like everyone else. To seek only praise and fear blame is
fruitless. The only blame with which we really need to be concerned
is that which is offered by the wise. If someone who lives impeccably
criticises us, it is appropriate to heed what they say. But if we
find that any criticism from anybody hurts us, then we need to look
deeper. Reflect that in blaming, people are sending out the pain that
they don't have the ability to contain; they express their hurt
outwardly by finding fault with others. When we have the mindfulness
and capacity to receive ourselves fully we are not inclined to blame
anyone, neither ourselves nor others.

\end{dhpRefl}

%% == 6 ==

\begin{dhpVerse}{328}
\label{dhp-328}
If you find a good companion,\\
of integrity and wisdom,\\
you will overcome all dangers\\
in joyous and caring company.
\end{dhpVerse}

\begin{dhpRefl}

Our mind, like water, takes on the shape of the vessel in which it is
contained. The Teacher is encouraging us to be mindful of the company
we keep. The Discourse on Great Blessings says, ``Avoid the company
of the foolish and associate yourself with the wise''. As we apply
skilful discrimination, we need to exercise care that we don't
confuse prejudice with wise discernment. Wise discernment is
compassionate and kind and is interested in protecting all beings
from harm.

\end{dhpRefl}

%% == 7 ==

\begin{dhpVerse}{8}
\label{dhp-8}
As a stormy wind cannot move\\
a mountain of rock,\\
so one who contemplates the reality of the body,\\
who develops faith and energy,\\
is unmoved by Māra.
\end{dhpVerse}

\begin{dhpRefl}

Māra is the manifestation of unenlightenment; the driving force of
denial. Māra is the avoidance of reality and is expressed as
compulsiveness, insensitivity, resentment. Our monastery in rural
Northumberland is sometimes buffeted by winds that can be frightening
but they are nothing compared to the threat posed by the forces of
Mara. If we are to withstand the onslaught of our heedless habits, we
must firmly establish our contemplation in an awareness of the body,
in faith and in energy. It is the ability and willingness to come
back, over and over again, to our body-based practice, remembering to
live in ways that lead to selfless confidence that sustains our
interest in discovering truth. This has the power to make us
unshakable.

\end{dhpRefl}

%% == 8 ==

\begin{dhpVerse}{104-105}
\label{dhp-104}\label{dhp-105}
Self-mastery is the supreme victory --\\
much more to be valued\\
than winning control over others.\\
It is a victory that no other being whatsoever\\
can distort or take away.
\end{dhpVerse}

\begin{dhpRefl}

Unshakably established in the centre of our lives, nothing can
unsettle us or generate causes for suffering. This is fearlessness.
The Buddha's teachings point to that which blocks fearlessness:
habitual self-seeking, deluded desire. Shining the light of awareness
on this very activity of desire -- not judging or interfering with it
in any way -- we learn to see that each small moment of self-mastery
leads towards the victory of selflessness.

\end{dhpRefl}

%% == 9 ==

\begin{dhpVerse}{268-269}
\label{dhp-268}\label{dhp-269}
Silence does not denote profundity\\
if you are ignorant and untrained.\\
Like one holding scales,\\
a sage weighs things up,\\
wholesome and unwholesome,\\
and comes to know both the inner and outer worlds.\\
Therefore the sage is called wise.
\end{dhpVerse}

\begin{dhpRefl}

The Buddha spoke of the contentment and benefit that can come from
living in quiet and beautiful places. Limiting sense stimuli can
assist us on our path to freedom from ignorance. However he didn't
mean for us to then take a position against the sensory world. Ajahn
Chah often said: ``If you can't practice in the city, you can't
practice in the forest''. And he'd also say: ``If you can't practice
when you are sick, you can't practice when you are healthy''. In
other words, everything is practice; including the feeling that we
can't practice with `this'. It is wisdom that recognises this truth.

\end{dhpRefl}

%% == 10 ==

\begin{dhpVerse}{172}
\label{dhp-172}
There are those who awaken from heedlessness.\\
They bring light into the world\\
like the moon\\
emerging from clouds.
\end{dhpVerse}

\begin{dhpRefl}

On the full moon day of the month of Asalha, over two thousand five
hundred years ago, the Buddha first revealed the Four Noble Truths.
The hearts of those who heard these teachings were filled with the
joy of perfect understanding, as cool and brilliant as a moon
emerging from a covering of clouds. We assume the world and our
suffering to be more substantial than it really is. But if we turn
our attention towards the true causes of suffering -- desire rooted
in ignorance - these Four Ennobling Truths generate a radiance that
disperses the clouds of delusion and dissolves the world of
suffering. Our contemplations awaken us from heedlessness, moment by
moment we contribute to the continued turning of this Dhamma Wheel.

\end{dhpRefl}

%% == 11 ==

\begin{dhpVerse}{78}
\label{dhp-78}
Do not seek the company of misguided friends;\\
beware of degenerate companions.\\
Seek and enjoy the company of well-guided friends,\\
those who support insight.
\end{dhpVerse}

\begin{dhpRefl}

How do we hold our friendships and how do our friends hold us? We
want to protect and nurture true friendship with those who support
our aspiration to live in truth. We genuinely appreciate such
companions; we don't wait until we get into difficulties before
letting them know what they mean to us. Good friendship can be
cultivated. As we bring mindfulness to our relationships with others
let's also consider how friendly we are being with ourselves.

\end{dhpRefl}

%% == 12 ==

\begin{dhpVerse}{145}
\label{dhp-145}
Those who build canals\\
channel the flow of water.\\
Arrowsmiths make arrows.\\
Woodworkers craft wood.\\
Those committed to goodness\\
tame themselves.
\end{dhpVerse}

\begin{dhpRefl}

Crafts-people work their crafts. The endeavour of taming the heart's
unruly nature -- there's an art in that. Outwardly we may find
ourselves in confusing circumstances but inwardly we remember that
cultivating skilfulness is our primary task. With careful, constant
observation we learn to recognise the heart's unruliness. `If only
the weather was warmer!' `If only the weather was cooler!' We don't
reactively judge what arises. Our passionate demanding that
conditions be other than how they are is simply seen for what it is.
There's an increased daring and willingness to offer ourselves into
whatever it is we are doing right now; whole-heartedly, whole-bodily.
As we progress, the task becomes easier; gratitude manifests, even
though the work is challenging.

\end{dhpRefl}

%% == 13 ==

\begin{dhpVerse}{146}
\label{dhp-146}
Why is there laughter?\\
Why is there joy\\
when the world is on fire?\\
Since you are clouded in darkness\\
should you not seek the light?
\end{dhpVerse}

\begin{dhpRefl}

It is usually when we get burned by life that we turn to the
Teachings. It can come as a relief to find that in our efforts to be
free from darkness we are accompanied by many millions of other human
beings. Suffering is the nature of unawakened humanity. ``Don't feel
bad if you are suffering. Everyone suffers,'' Ajahn Chah would tell
us. Before his enlightenment the Buddha-to-be suffered too. The
difference is enlightened beings know that suffering is not an
obligation -- it is only one of the options available in the human
realm. There is also the possibility of dwelling in the light of
non-suffering.

\end{dhpRefl}

%% == 14 ==

\begin{dhpVerse}{239}
\label{dhp-239}
Gradually, gradually,\\
a moment at a time,\\
the wise remove their own impurities\\
as a goldsmith removes the dross.
\end{dhpVerse}

\begin{dhpRefl}

No amount of wishing things were otherwise gives us what we long for.
We want the pure gold of pristine awareness so we need to enter the
fires of purification. This verse instructs us on how to watch over
the burning: too much heat -- we are trying too hard -- enduring
heedlessly we get hurt in our practice. Not enough heat -- shying
away from difficulties -- following preferences for comfort and ease
there is no improvement in our practice. We just become more foolish
as the years go by. Our habits are the dross and with gradual
fine-tuning of our effort we learn letting go. The aim of all this
work is the realisation of the state of luminous awareness. We then
have something inherently valuable to share with others.

\end{dhpRefl}

%% == 15 ==

\begin{dhpVerse}{178}
\label{dhp-178}
Better than ruling the whole world,\\
better than going to heaven,\\
better than lordship over the universe,\\
is an irreversible commitment to the Way.
\end{dhpVerse}

\begin{dhpRefl}

Unconditioned freedom: a quality of being not dependent on any
condition whatsoever. Whatever the circumstances, fortunate or
unfortunate, the heart remains at ease, radiant, clear-seeing,
sensitive and strong. It is a commitment that is irreversible,
unshakable and real; beyond all of deluded ego's obsessions. To
arrive at this level of resolve requires constant observation of our
old habits, for example: the fondness of being in control of
everything, addiction to passing pleasures, obsession with power. We
work with what we have. Each time we find ourselves distracted we
refocus our commitment to the Way.

\end{dhpRefl}

%% == 16 ==

\begin{dhpVerse}{76}
\label{dhp-76}
Only blessings\ldots

MISSING
\end{dhpVerse}

\begin{dhpRefl}
There are two ways\ldots

MISSING
\end{dhpRefl}

%% == 17 ==

\begin{dhpVerse}{1}
\label{dhp-1}
All states of being are determined by mind.\\
It is mind that leads the way.\\
Just as the wheel of the oxcart follows\\
the hoof-print of the animal that draws it,\\
so suffering will surely follow\\
when we speak or act impulsively\\
from an impure state of mind.
\end{dhpVerse}

\begin{dhpRefl}

Thinking too much can complicate spiritual practice; not thinking
enough can restrict us. Here we are given an example of
wisely-directed thinking. It helps us to see that we are not a victim
of circumstance -- to understand that the intention behind our
actions of body and speech determines our state of being -- it puts
us in the position where we have the power to bring about real
change. Accepting this teaching as our field of investigation we
appreciate the value of exercising mindful restraint and find new
confidence and ability. Our heart/mind leads the way.

\end{dhpRefl}

%% == 18 ==

\begin{dhpVerse}{118}
\label{dhp-118}
Having performed a wholesome act\\
it is good to repeat it.\\
Enjoy the pleasure of its memory.\\
The fruit of goodness is contentment.
\end{dhpVerse}

\begin{dhpRefl}

The Dhamma encourages us to bring to heart the memory of wholesome
acts performed by body, speech or mind and to dwell in the sense of
gladness that arises with that recollection. We are not at risk of
becoming overly pleased with ourselves if mindfulness is present. We
can similarly contemplate our mistakes and shortcomings without
losing sight of our goodness if we have developed wise reflection.
That we can observe our habits shows that we are much more than those
habits. What is it that observes? This is our refuge - awareness -
the way out of suffering. With this very awareness we can freely
dwell in the joy of recollecting goodness. And what a relief to find
that we can make mistakes and learn from them.

\end{dhpRefl}

%% == 19 ==

% @@

\begin{dhpVerse}{348}
\label{dhp-348}
Let go of that which is in front,\\
let go of that which has already gone,\\
and let go of in-between.\\
With a heart that takes hold nowhere\\
you arrive at the place beyond all suffering.
\end{dhpVerse}

\begin{dhpRefl}

We can comfortably lose ourselves in ideas of the future. We can
comfortably dwell in memories of the past. And we readily get lost in
experiences here and now. When we get lost we suffer. But suffering
is not ultimate; there are causes for suffering and there is freedom
from them. To arrive at freedom requires that we are willing to leave
our familiar abidings. We hear the instruction to `let go' and it can
sound like we are being told to get rid of something or that we are
wrong for being the way we are. But if we recognise the power of
right mindfulness, we see letting go just happens. In this verse the
Buddha meets us in the place of our suffering and directs us towards
an open door out of it.

\end{dhpRefl}

%% == 20 ==

\begin{dhpVerse}{387}
\label{dhp-387}
The sun shines by day,\\
the moon shines by night.\\
But, both all day and all night\\
the Buddha shines\\
in glorious splendour.
\end{dhpVerse}

\begin{dhpRefl}

Whenever there is well-established mindfulness, there is beauty,
clarity and the possibility for deepening understanding. If there is
constant mindfulness, there is constant clarity. All day and all
night the heart of mindfulness shines in glorious splendour.

\end{dhpRefl}

%% == 21 ==

\begin{dhpVerse}{122}
\label{dhp-122}
Do not ignore the effect of right action\\
saying, ``This will come to nothing.''\\
Just as by the gradual fall of raindrops\\
the water jar is filled,\\
so in time the wise become replete with good.
\end{dhpVerse}

\begin{dhpRefl}

`Bigger is better'. `The more the merrier'. Little by little such
perceptions can be wisely questioned. Trust that every small moment
of mindfulness matters. Every small effort to remember to come back
and with judgment-free awareness to begin again makes a difference.
No such effort is ever wasted. One day we discover we are simply
unimpressed by that which previously held us. Instead of reacting to
something that in the past would upset us, letting go happens. Wisdom
knows the way of true goodness.

\end{dhpRefl}

%% == 22 ==

\begin{dhpVerse}{134}
\label{dhp-134}
If spoken to harshly,\\
make yourself as silent as a cracked gong;\\
non-retaliation is a sign of freedom.
\end{dhpVerse}

\begin{dhpRefl}

When we receive unjustified criticism it can be hard to restrain the
up-thrusts of passion. Pushing strong feelings down into unawareness
isn't helpful. Practice means finding the space within ourselves to
feel what we feel, without `becoming' those feelings. It is a special
skill. Watch out for any voice preaching at you: ``You shouldn't be
this way, you should know better by now.'' We acknowledge the fact of
how it is in this moment. Accept the present reality into awareness,
knowing it as it is; not indulging, not pushing away. Now the energy
of our passions can fuel the process of purification, burning out the
pollutions rather than burning us out with self-criticism.

\end{dhpRefl}

%% == 23 ==

\begin{dhpVerse}{130}
\label{dhp-130}
Having empathy for others\\
one sees that all beings\\
love life and fear death.\\
Knowing this,\\
one does not attack or cause attack.
\end{dhpVerse}

\begin{dhpRefl}

By being open to others we bring benefit to ourselves. In studying
our own hearts and minds we generate benefit for others. Dhamma
teaches us to practice on both levels. On one level we cultivate
insight that undoes deeply held views about `self' and `others'. This
is our formal practice. Meanwhile our daily-life practice calls us to
accept the relative validity of views like `self' and `others'; to
engage them in the cultivation of compassion and kindness. Empathy is
a companion of insight and helps keep practice balanced.

\end{dhpRefl}

%% == 24 ==

\begin{dhpVerse}{262-263}
\label{dhp-262}\label{dhp-263}
Those who are envious, stingy and manipulative,\\
remain unappealing despite good looks\\
and eloquent speech.\\
But those who have freed themselves\\
from their faults\\
and arrived at wisdom\\
are attractive indeed.
\end{dhpVerse}

\begin{dhpRefl}

We need to remind ourselves that the way things appear to be is not
necessarily the way things `are'. If we listen from the heart to
someone as they speak, instead of listening only with our ears, we
will hear something quite different. When we view others from a quiet
centred place deep within, we see much more than that which our eyes
see.

\end{dhpRefl}

%% == 25 ==

\begin{dhpVerse}{162}
\label{dhp-162}
Committed evil doers behave toward themselves\\
like their own worst enemies.\\
They are like creepers that strangle the trees\\
which support them.
\end{dhpVerse}

\begin{dhpRefl}

This verse refers to a monk who tried three times to kill the Buddha
and eventually his evil actions contributed to his own death. When we
betray our heart's commitment to reality, slowly but surely, we die
away from the light of truth and sink into darkness. Creepers climb
handsome, mature trees and sometimes strangle them to death. We can
go for refuge to the Buddha yet still be taken over by up-thrusts of
rage. Hours, days or even years go by as we justify our hurtful
actions of body or speech. When we come to see the truth of our
actions a wholesome sense of remorse arises; we genuinely wish to
desist. Right action is the natural consequence.

\end{dhpRefl}

%% == 26 ==

\begin{dhpVerse}{184}
\label{dhp-184}
A renunciate does not oppress anyone.\\
Patient endurance is the ultimate asceticism.\\
Profound liberation, say the Buddhas,\\
is the supreme goal.
\end{dhpVerse}

\begin{dhpRefl}

As we progress on our path towards the goal, the Buddha is
encouraging us not to forget how we relate with others. The
cultivation of metta teaches us to include all beings in our heart of
kindness. It sometimes happens that we feel able to be kind towards
others but not towards ourselves. If we find this a struggle, we show
infinite kindness towards the struggler, ourselves. Layers of
self-judging, self-loathing, self-rejecting require us to exercise
patient endurance; not to be mistaken with bitter endurance. Patience
is an essential characteristic for those travelling on the path. And
there is absolutely no way this essential virtue can be cultivated
when we are having a good time. So when we are having a bad time,
reflect that this is the perfect place, the only place, to develop
this profound quality.

\end{dhpRefl}

%% == 27 ==

\begin{dhpVerse}{6}
\label{dhp-6}
Those who are contentious\\
have forgotten that we all die;\\
for the wise, who reflect on this fact,\\
there are no quarrels.
\end{dhpVerse}

\begin{dhpRefl}

It is OK to think about death and to talk about the fact that we all
will die. Indeed it is wise to do so. If we deny the awareness of our
mortality, the fear of it remains hidden underground. With this
avoidance of reality comes diminishing aliveness and increased
confusion. Cultivating mindfulness around the subject of death
(maranasati) we discover that to the degree we admit reality there is
a corresponding increase in clarity and contentment. This is the
contentment that leads to meaningful peace.

\end{dhpRefl}

%% == 28 ==

\begin{dhpVerse}{2}
\label{dhp-2}
All states of being are determined by mind.\\
It is mind that leads the way.\\
As surely as our shadow never leaves us,\\
so well-being will follow when we speak or act\\
with a pure state of mind.
\end{dhpVerse}

\begin{dhpRefl}

Have you ever tried to run away from your shadow? As hard as you
might try it never leaves you. Any action by body, speech or mind
that is done with pure motivation, will indeed lead to increased
well-being. This is worth remembering if we intend to do something
good and are about to change our minds, thinking it's not worth the
effort. However small and apparently insignificant the act might be,
happiness will follow. It is worth doing.

\end{dhpRefl}

%% == 29 ==

\begin{dhpVerse}{3-4}
\label{dhp-3}\label{dhp-4}
When we hold fast to such thoughts as,\\
``They abused me, mistreated me,\\
molested me, robbed me,''\\
we keep hatred alive.

If we thoroughly release ourselves\\
from such thoughts as,\\
``They abused me, mistreated me,\\
molested me, robbed me,''\\
hatred is vanquished.
\end{dhpVerse}

\begin{dhpRefl}

In various ways we all suffer injustice in our lives. Sometimes the
pain runs deep and can last for years. Dhamma teachings emphasise not
so much the pain but our relationship to it. As long as we are
possessed by hatred and resistance, intelligence is compromised.
Although action might be called for, if our heart is not free from
hatred, we can't know what right action would be. The Buddha
advocates releasing ourselves from thoughts of hatred. It takes
strength, patience and determination to let go. We are not letting go
because someone else told us we should. We are letting go because we
understand the consequences of hanging on.

\end{dhpRefl}

%% == 30 ==

\begin{dhpVerse}{377}
\label{dhp-377}
As old flowers fall\\
from a jasmine plant\\
let lust and hatred\\
fall away.
\end{dhpVerse}

\begin{dhpRefl}

Ajahn Chah was in London, staying at the Hampstead Vihara. The monks
were troubled by the noise that was coming from the pub across the
road. Ajahn Chah told them that the cause of suffering was their
sending attention out to trouble the sound. Sound itself is just so.
Suffering only arises when we `go out' and add something extra.
Seeing our part in creating problems, a shift in the way we view
struggles takes place. Instead of blaming we simply `see' what we are
doing, in the moment. Let's not get into a fight with hatred;
exercising careful restraint and wise reflection, we let it `fall
away'. Initially we see this only after we have reacted and created
suffering. With practice we catch it sooner. One day, we will catch
ourselves just as we are about to create the problem.

\end{dhpRefl}

%% == 31 ==

\begin{dhpVerse}{401}
\label{dhp-401}
As water slides from a lotus leaf,\\
so sensual pleasures\\
do not cling\\
to a great being.
\end{dhpVerse}

\begin{dhpRefl}

A great being is great because he or she is free from obstructions in
their relationship with life. We are not so great because we get
caught in feelings and make a problem out of life. We create
obstructions by the way that we deal with the eight worldly dhammas:
praise and blame, gain and loss, pleasure and suffering, popularity
and insignificance. Out of delusion we relate to these worldly winds
heedlessly - indulging in what we like and resisting what we don't.
Wisdom on the other hand simply sees the reality of the sensory
world. It knows the space within which all experiences arise and
cease. Such knowing means a great being doesn't even have to try to
let go; all inclination to cling automatically falls away. He or she
experiences sensual pleasure but adds nothing to it and takes nothing
away.

\end{dhpRefl}

%% == 32 ==

\begin{dhpVerse}{5}
\label{dhp-5}
Never by hatred is hatred conquered,\\
but by readiness to love alone.\\
This is eternal law.
\end{dhpVerse}

\begin{dhpRefl}

To conquer hatred `by love alone' can seem a lofty ideal. We might
judge ourselves (and in so doing tend to excuse ourselves) as being
too limited -- or just dismiss the notion as not realistic. Surely
part of us does say, `yes'. Another part of us might say, `Yes, but
what if\ldots{}' For the Buddha there are no qualifications;
hatred simply never works. What is this word `love'? Generating an
unqualified receptivity? And the perceptions we have of those who
might trigger hatred within us; where do those perceptions exist?
Potentially loving-kindness is capable of receiving all of this into
awareness, fully, wholly, even warmly. This is eternal law.

\end{dhpRefl}

%% == 33 ==

\begin{dhpVerse}{142}
\label{dhp-142}
Flamboyant outer appearance\\
does not in itself constitute\\
an obstruction to freedom.\\
Having a heart at peace, pure, contained,\\
awake and blameless,\\
distinguishes a noble being.
\end{dhpVerse}

\begin{dhpRefl}

Outer forms are not what really matter. The Buddha always pointed to
the heart, the place we need to focus. He emphasized this because we
easily forget, becoming too concerned with how things appear. This
verse is about an inebriated householder shocked by massive
disappointment into despair. The Buddha's teachings, pointing
directly at that which really matters, transformed his despair;
instead of falling apart, he realised perfect unconditioned peace.
When the focus of our practice is merely on forms and appearances,
spirit is compromised. Straining too much for the ultimate
explanation of the five precepts, for example, may stop us from
seeing the intention behind the action. The form of the precepts is
there so we become mindful of our motivation. Holding the precepts
correctly, there is a chance they will serve their true purpose.
Correct living is noble being.

\end{dhpRefl}

%% == 34 ==

\begin{dhpVerse}{95}
\label{dhp-95}
There are those who discover\\
they can leave behind confused reactions\\
and become patient as the earth;\\
unmoved by anger,\\
unshaken as a pillar,\\
unperturbed as a clear and quiet pool.
\end{dhpVerse}

\begin{dhpRefl}

The Buddha lived in this world as we do. And despite all the turmoil
he realised a state of imperturbability - to `become patient as the
earth'. Whatever is poured on it, burnt on it, or done to it, the
earth just goes on being what it is, doing what it does. Patient
endurance is not a weak option; it is strengthening and kind. With
patient endurance we discover the ability to allow our present
experience to be here, just as it is, until we have learned what we
need to learn.

\end{dhpRefl}

%% == 35 ==

\begin{dhpVerse}{256}
\label{dhp-256}
Making an arbitrary decision\\
does not amount to justice.\\
Having considered arguments\\
for and against,\\
the wise decide the case.
\end{dhpVerse}

\begin{dhpRefl}

You are under pressure to make a decision. Is it possible to remain
cool and calm when others want you to decide in their favour? Can
you remain free from bias and arrive at a just decision? How do we
hold our views? Having a strong opinion can feel great; it can appear
as confidence. But such is the nature of fundamentalism; so too is
providing simplistic answers to complex questions. Rigid views and
simplistic solutions are not aspects of a spacious mind; a mind that
can consider all aspects of a dilemma. It usually takes time to
arrive at a balanced, thoroughly considered view. It also requires an
ability to listen from a place of inner quiet. If we are mentally
preparing our rebuttal we are not really listening.

\end{dhpRefl}

%% == 36 ==

\begin{dhpVerse}{120}
\label{dhp-120}
Even those who live wholesome lives\\
can experience suffering so long as their acts\\
have not yet borne direct fruits.\\
However, when the fruits of their actions ripen\\
the joyful consequences cannot be avoided.
\end{dhpVerse}

\begin{dhpRefl}

Trusting in true principles is not always easy. Pursuing popular
opinion is not always wise. Not being dissuaded by doubts or mere
speculation, we see how our deepest questions about reality lead us
to insight. Trust is a force for transformation, so too is patience.
It is almost unimaginable how an acorn can grow into an oak tree, but
it does. We need trust. We need patience. Like trees, these forces
for transformation can be cultivated and inspire us towards that
which is inwardly great and inherently free of all suffering.

\end{dhpRefl}

%% == 37 ==

\begin{dhpVerse}{91}
\label{dhp-91}
Alert to the needs of the journey,\\
those on the path of awareness,\\
like swans, glide on,\\
leaving behind their former resting places.
\end{dhpVerse}

\begin{dhpRefl}

It is time to move on. We might know this on one level yet another
part of us holds back. If we attempt to let go because we think we
should -- instead of discovering renewed inspiration -- there will be
resistance. But encountering disappointment is OK too if we are alert
to our heart's true needs. Our deepest need is to be free from
ignorance. There is no place we can truly settle until we have fully
realised who and what we are. This path of awareness has the power to
transform absolutely everything into understanding. Swans glide on --
no resistance.

\end{dhpRefl}

%% == 38 ==

\begin{dhpVerse}{223}
\label{dhp-223}
Transform anger with kindness\\
and evil with good,\\
meanness with generosity\\
and deceit with integrity.
\end{dhpVerse}

\begin{dhpRefl}

If we are cold, we find a way to get warm -- we don't subject
ourselves to more cold. If we are hungry we eat -- we don't further
deprive ourselves of food. If we are angry, we don't fight it with
rage -- we try to be kind to this being that is suffering from anger.
If we witness evil actions, we generate sincere goodness and restrain
any impulse to reject the evil-doers with mere judgment. To people
who believe self-centred meanness is the path to contentment we give
generosity. And with those who are deceitful we speak truth. This may
not be easy. This is the way of transformation.

\end{dhpRefl}

%% == 39 ==

\begin{dhpVerse}{258}
\label{dhp-258}
Those who speak much\\
are not necessarily possessed of wisdom.\\
The wise can be seen to be at peace with life\\
and free from all enmity and fear.
\end{dhpVerse}

\begin{dhpRefl}

Struggling to become peaceful and wise often stirs things up while
carefully allowing the activity of heart and mind to cease naturally
calms things down. Wisdom is not discovered by following habits of
seeking that which we like. When pain and disappointment come to us
we can learn to meet them with a quality of willingness and awareness
that leads to understanding. When happiness and delight come we can
be refreshed and renewed by our experience without becoming lost in
them. Enmity and fear are not necessarily obstacles. They can be our
teachers pointing to a more spacious and wise way of living.

\end{dhpRefl}

%% == 40 ==

\begin{dhpVerse}{82}
\label{dhp-82}
On hearing true teachings,\\
the hearts of those who are receptive\\
become serene, like a lake,\\
deep, clear and still.
\end{dhpVerse}

\begin{dhpRefl}

We might think, `If the Buddha were here today to teach me, I too
could get enlightened.' However when the Buddha was around there were
many who met him, heard him, lived with him and yet didn't recognise
him or the Truth to which he was pointing. At the end of his life one
of the monks asked the Buddha who would take his place when he was
gone, to which he replied that the Dhamma would take his place. On an
earlier occasion he had taught that seeing the Dhamma was the same as
seeing the Buddha. Listening to Dhamma is attending fully to and
contemplating wisely that which is happening right now. So we don't
need to feel bad because we can't listen to a talk directly from the
Buddha. We learn to be more receptive to life.

\end{dhpRefl}

%% == 41 ==

\begin{dhpVerse}{173}
\label{dhp-173}
One who transforms old and heedless ways\\
into fresh and wholesome acts\\
brings light into the world\\
like the moon freed from clouds.
\end{dhpVerse}

\begin{dhpRefl}

The darkness of the world comes from habits of avoidance and
indulgence. Wholesome acts uncover the light that is already here.
Paying attention to old worn-out ways in the right manner at the
right time with here-and-now, judgment-free, body-mind awareness
leads to transformation. We don't have to get rid of anything. We
don't have any bad stuff. Painful memories, painful sensations do not
have to become suffering. With right attention everything leads to
increased light.

\end{dhpRefl}

%% == 42 ==

\begin{dhpVerse}{186-187}
\label{dhp-186}\label{dhp-187}
Not in great wealth is there contentment,\\
nor in sensual pleasure,\\
gross or refined.\\
But in the extinction of craving\\
is joy to be found by a disciple of the Buddha.
\end{dhpVerse}

\begin{dhpRefl}

Is what I am doing with my life giving me what I am looking for? We
do succeed from time to time in gratifying our wishes but lasting
happiness comes only from being freed of the irritation of wanting.
If we ask ourselves the right questions at the right time, a
radically different view is revealed: the way of contentment draws us
inwards, not outwards. Rather than following an impulse to want more,
we can look directly at the wanting itself. And for a moment the
sting of craving ceases. We have learned a little more about what it
means to be a disciple of the Buddha.

\end{dhpRefl}

%% == 43 ==

\begin{dhpVerse}{14}
\label{dhp-14}
As rain cannot penetrate\\
a well-thatched roof,\\
so the passions\\
cannot enter a well-trained heart.
\end{dhpVerse}

\begin{dhpRefl}

We act to protect ourselves from the elements of wind and rain. With
wise reflection we act to protect ourselves from the wild passions.
This is heart-training. Skilfully addressing the damage caused by
greed and ill will is an important aspect of Right Action. It is not
an avoidance of the passions; neither is it indulging in them.
Between these two we seek the middle way.

\end{dhpRefl}

%% == 44 ==

\begin{dhpVerse}{143}
\label{dhp-143}
A well-trained horse\\
gives no cause for restraint.\\
Rare are those beings who,\\
through modesty and discipline,\\
give no cause for rebuke.
\end{dhpVerse}

\begin{dhpRefl}

Is it possible to have too much modesty? It is possible to have the
wrong kind of modesty; when, for example, refraining from blatant
greed is being manipulative. Modesty, frugality and discipline: words
like these can make us feel uneasy. Yet certainly there are timeless
principles encoded in them. When we have true modesty we look for the
`right amount' of things. We seek the difference between settling for
`good enough' and being too timid to excel. With right discipline we
are focussed on the task at hand without compromising sensitivity.
One skilled in right discipline can say no when needed, not out of
judgement or mere preference, but as someone who cares.

\end{dhpRefl}

%% == 45 ==

\begin{dhpVerse}{49}
\label{dhp-49}
As a bee gathering nectar\\
does not harm or disturb\\
the colour and fragrance of the flower,\\
so do the wise move\\
through the world.
\end{dhpVerse}

\begin{dhpRefl}

The world is our village. Can we move through our world without
disturbing that which is beautiful about it? Sages and renunciates
might live on alms-food gathered from the village but most people
need to use money. There is nothing inherently wrong about money. It
is a symbol for the energy with which we interact in our world. The
Buddha's guidelines on cultivating Right Livelihood indicate harmless
and considerate ways of stewarding this energy. When we forget
ourselves we focus too much on how good it will be when we get what
we want, not attending to the way we go about getting it. Care is
needed both in what we do and how we do it.

\end{dhpRefl}

%% == 46 ==

\begin{dhpVerse}{217}
\label{dhp-217}
Naturally loved are those\\
who live with right action\\
and have found the Way,\\
and through insight\\
have become established in the truth.
\end{dhpVerse}

\begin{dhpRefl}

When actions of body and speech are an expression of inner ease, we
find the realm of relationships more rewarding. It's not by merely
following our wish to be liked or any other gratification of desire
that we live with lasting happiness; rather it is by appreciating
where the true causes for contentment lie. Even wholesome wishes if
held too tightly can lead to discontentment. It's clinging that's the
problem. When we see clearly our habits of clinging, at that time and
in that place we also see how we don't have to cling; we find the way
out.

\end{dhpRefl}

%% == 47 ==

\begin{dhpVerse}{20}
\label{dhp-20}
Knowing only a little about Dhamma\\
but wholeheartedly according with it,\\
transforming the passions\\
of greed, hatred and delusion,\\
releasing all attachments to here and hereafter,\\
one will indeed experience for oneself\\
the benefits of walking the Way.
\end{dhpVerse}

\begin{dhpRefl}

It doesn't matter that we don't know all there is to know about
Dhamma. It is how we apply the teachings that counts. Are we living
wholeheartedly according with what we have learned? This is a more
useful line of enquiry than, `When can I go on another retreat?' or
`If I could only knuckle down to more study?' Such questions are OK
if they lead us to releasing attachments but not if they are an
expression of our addiction to always wanting or becoming more. If
what is needed is a subtle shift of attention to simply being more
mindful in the present moment, even striving hard for profound
insights obstructs our progress. A moment of recognising this truth
and we experience for ourselves the benefit of walking the Way.

\end{dhpRefl}

%% == 48 ==

\begin{dhpVerse}{341}
\label{dhp-341}
Beings naturally experience pleasure;\\
but when pleasure is contaminated with craving,\\
not releasing it creates frustration\\
and tedious suffering follows.
\end{dhpVerse}

\begin{dhpRefl}

With the right quality of mindfulness at the right moment, we can see
why and how attaching to joy and sorrow perpetuates suffering. We
become interested in how to experience pleasure without making it
`my' pleasure or creating an `I' that is having a good time. Wise
reflection shows us that when attachment is taken out of the
experience happiness isn't diminished -- in fact it is enhanced. When
we are accurately aware, intelligence functions in the service of
insight. If we appreciate how grasping at the pleasure spoils the
beauty, an ease of being emerges.

\end{dhpRefl}

%% == 49 ==

\begin{dhpVerse}{188-189}
\label{dhp-188}\label{dhp-189}
To many places beings withdraw\\
to escape from fear:\\
to mountains, forests, parklands\\
and gardens; sacred places as well.\\
But none of these places offer true refuge,\\
none of them can free us from fear.
\end{dhpVerse}

\begin{dhpRefl}

It is hard to feel afraid without thinking something is going wrong.
We readily react by judging ourselves and others, in an attempt to
escape the pain of fear. It doesn't work -- neither does running off
into the wilds. Even sacred places are deemed to fail us if we are
motivated by a wish to escape. Turning to our refuge in Dhamma
however can trigger an interest in understanding fear and learning
from fear. Can we experience the fear sensation without `becoming'
afraid? Fear is still fear but it is perceived from an expanded, less
cramped and less threatened awareness. We can even begin to see that
fear too is `just so'. A non-judgemental, whole body-mind
acknowledgement of the condition of fear, here and now, can transform
our pain into freedom. Willingness to meet ourselves where we find
ourselves is the way.

\end{dhpRefl}

%% == 50 ==

\begin{dhpVerse}{193}
\label{dhp-193}
It is hard to find\\
a being of great wisdom;\\
rare are the places in which they are born.\\
Those who surround them when they appear\\
know good fortune indeed.
\end{dhpVerse}

\begin{dhpRefl}

Associating with wise beings is certainly a blessing but it is not
always easy. Their openness can contrast painfully with our
contractedness. However, wise beings are also compassionate so it is
easier to take the medicine. Eventually we feel grateful. After his
enlightenment the Buddha felt so much gratitude for his previous
teachers that one of the first things he did was try to find them and
pass on what he had discovered. We too can reflect with gratitude on
all those who have helped us take steps on this path. Dwelling on
gratitude it is heart nourishment; it aligns us with the wise.

\end{dhpRefl}

%% == 51 ==

\begin{dhpVerse}{179}
\label{dhp-179}
The Buddha's perfection is complete;\\
there is no more work to be done.\\
No measure is there for his wisdom;\\
no limits are there to be found.\\
In what way could he be distracted from truth?
\end{dhpVerse}

\begin{dhpRefl}

There is a real reality that can be known; the Buddha knew it. It is
possible to awaken beyond all the feelings of limitation that lead to
disappointment and regret. Skilfully trusting in the Buddha's
enlightenment establishes an inner frame of reference. When we find
ourselves disoriented by life we realign ourselves by engaging this
trust. Trust is an inner structure, like the sails on a yacht; when
rightly set they capture the wind which powers the vessel onwards.

\end{dhpRefl}

%% == 52 ==

\begin{dhpVerse}{153-154}
\label{dhp-153}\label{dhp-154}
For many lives I have wandered\\
looking for, but not finding,\\
the house-builder who caused my suffering.\\
But now you are seen and\\
you shall build no more.\\
Your rafters are dislodged and\\
the ridge-pole is broken.\\
All craving is ended;\\
my heart is as one\\
with the unmade.
\end{dhpVerse}

\begin{dhpRefl}

The Buddha's final and complete realisation was to discover that he
had been believing in something that was not true. He had been
deceived by what he calls the house-builder. The houses are the
structures of mind; the `me' and the `mine' we take so seriously. `It
is me that is longing.' `It is me that is feeling disappointed.'  `It
is my mood, my body, my mind.'  He saw clearly how these impressions
had been constructed by the habits of craving. Having penetrated this
with profound insight, the process was understood; the mainstay of
the house, the ridge-pole, was broken and suffering had ended.
Wandering around hoping for a solution to his struggles was over.
From here on, he would abide at ease in the state prior to all
arising and ceasing; the unmade, the undying reality.

\end{dhpRefl}

